---
title: "Pollution Analysis (01/25/2023 - 02/01/2023)"
format:
  html:
    page-layout: "full"
    code-fold: false
    code-tools:
      source: https://github.com/dmolitor/air-pollution
    self-contained: true
    theme: journal
    toc: false
execute: 
  cache: false
highlight-style: dracula
---

## Data-cleaning preamble

Everything here is calculated using a rather small subset of the original set of
$3172$ monitors spanning 1999-2020.
To get to this subset I took the following steps:

1) For all co-located monitors, I dropped all but the monitor that was **first**
installed at that location. Drops it down to $1967$ monitors (!!! big drop).
2) Then I create a 5-mile buffer (slightly different from last week at 10-mile
level) around each of these monitors and drop any monitors whose radius overlaps
with another monitor's radius. After this there are $595$ monitors (!!! again).

Finally, I was a little worried about the results being skewed by the fact that
the distribution of start dates for these monitors was heavily biased towards
1999-2000, as shown below.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
library(sf)
source(here::here("src/utils.R"))

all_monitors <- readRDS(here::here("data/monitor-pollution.Rds"))
all_monitors_dedup_site <- all_monitors |>
  dplyr::group_by(geometry) |>
  dplyr::mutate(first_monitor_start_date = min(monitor_start_date)) |>
  dplyr::filter(monitor_start_date == first_monitor_start_date) |>
  dplyr::slice_head(n = 1) |>
  dplyr::ungroup() |>
  dplyr::mutate(row_id = dplyr::cur_group_rows())

all_monitors_dedup_site_buffer <- all_monitors_dedup_site |>
  bufferize_points(buffer = 5, unit = "mi")

all_monitors_dedup_site_buffer$intersects_geom <- (
  lengths(sf::st_intersects(all_monitors_dedup_site_buffer)) > 1
)
all_monitors_dedup_site_buffer <- all_monitors_dedup_site_buffer |>
  dplyr::filter(!intersects_geom)
all_monitors_dedup_site <- all_monitors_dedup_site |>
  dplyr::filter(row_id %in% all_monitors_dedup_site_buffer$row_id)

ggplot2::ggplot(all_monitors_dedup_site) +
  ggplot2::geom_histogram(
    ggplot2::aes(x = monitor_start_date),
    color = "gray40",
    bins = 50
  ) +
  ggplot2::scale_x_date(date_breaks = "10 month", date_labels = "%B %Y") +
  ggplot2::labs(x = "Monitor start date", y = "Count") +
  theme_minimal_light() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, face = "bold"),
    axis.title = ggplot2::element_text(face = "bold"),
    panel.grid.minor.x = ggplot2::element_blank()
  )
```

<!--
```{r}
# Okay, not sure this is really a valid concern, but it worried me so I restricted
# the time range of this analysis to be dates after January 1, 2000. After this
# final restriction we have a subset of $265$ out of $3172$ original monitors 😬.
# The distribution of monitor start dates is now much more uniform-looking.
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
# all_monitors_dedup_site_buffer <- all_monitors_dedup_site_buffer |>
#   dplyr::filter(monitor_start_date >= as.Date("2001/01/01"))
all_monitors_dedup_site <- all_monitors_dedup_site |>
  dplyr::filter(row_id %in% all_monitors_dedup_site_buffer$row_id)

ggplot2::ggplot(all_monitors_dedup_site) +
  ggplot2::geom_histogram(
    ggplot2::aes(x = monitor_start_date),
    color = "gray40",
    bins = 50
  ) +
  ggplot2::scale_x_date(date_breaks = "10 month", date_labels = "%B %Y") +
  ggplot2::labs(x = "Monitor start date", y = "Count") +
  theme_minimal_light() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, face = "bold"),
    axis.title = ggplot2::element_text(face = "bold"),
    panel.grid.minor.x = ggplot2::element_blank()
  )
```
-->

But, I decided not to worry too much about this, and push onward 😬!
Of the remaining monitors, there's pretty good geographic dispersion. So it's
not like the only remaining monitors are clustered in a particular area.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
us_boundary <- tigris::states(progress_bar = FALSE, cb = TRUE) |>
  dplyr::filter(NAME %in% states())

ggplot2::ggplot(all_monitors_dedup_site, ggplot2::aes(color = state)) +
  ggplot2::geom_sf(data = us_boundary, alpha = 0, inherit.aes = FALSE) +
  ggplot2::geom_sf() +
  theme_minimal_light() +
  ggplot2::theme(legend.position = "none")
```

## Pollution trends for non-overlapping monitors

Okay, now that we have a bunch of monitors that are non-overlapping we can see
what the pre-post monitor installment trends look like.

### Raw PM2.5 values

#### Non de-trended PM2.5 values

For each monitor, I've calculated the mean/median PM2.5 for the 12 months before
and after monitor installment. The mean of individual monitor means is shown
below.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
monitors <- readRDS(
  here::here("data/monthly-pm2.5/monitor-raw-pollution.Rds")
) |> dplyr::filter(row_id %in% all_monitors_dedup_site_buffer$row_id)
pollution <- new_pollutionSeqCollection(
  monitors$pollution,
  summary_names = attr(monitors$pollution[[1]], "summary_names")
)
pm_trend <- dplyr::summarise(
  dplyr::group_by(enframe(pollution), event_seq),
  mean_pm = mean(Mean),
  median_pm = median(Median)
)

ggplot2::ggplot(pm_trend, ggplot2::aes(x = event_seq, y = mean_pm, group = 1)) +
  ggplot2::geom_line(color = "gray50") +
  ggplot2::geom_point(color = "palegreen4") +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "thistle4",
    linetype = "dashed"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = mean_pm),
    data = dplyr::filter(pm_trend, event_seq <= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = mean_pm),
    data = dplyr::filter(pm_trend, event_seq >= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(-12, 12, by = 2)
  ) +
  theme_minimal_light() +
  ggplot2::labs(
    x = "Months pre/post monitor start month",
    y = "Mean of monitor mean PM2.5"
  )
```

Similarly, the median of monitor median monthly values.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
ggplot2::ggplot(pm_trend, ggplot2::aes(x = event_seq, y = median_pm, group = 1)) +
  ggplot2::geom_line(color = "gray50") +
  ggplot2::geom_point(color = "palegreen4") +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "thistle4",
    linetype = "dashed"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = median_pm),
    data = dplyr::filter(pm_trend, event_seq <= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = median_pm),
    data = dplyr::filter(pm_trend, event_seq >= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(-12, 12, by = 2)
  ) +
  theme_minimal_light() +
  ggplot2::labs(
    x = "Months pre/post monitor start month",
    y = "Median of monitor median PM2.5"
  )
```

#### PM2.5 values with seasonality removed (via STL)

For each monitor, I've calculated the mean PM2.5 for the 12 months before
and after monitor installment. Then I apply STL to this two-year time series of
pollution values and remove the seasonal component. This feels a little hokey ...
idk, but also feels somewhat reasonable; I'm torn. The mean of de-seasonalized
individual monitor means is shown below.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
monitors <- readRDS(
  here::here("data/monthly-pm2.5/monitor-detrended-pollution.Rds")
) |> dplyr::filter(row_id %in% all_monitors_dedup_site_buffer$row_id)
pollution <- new_pollutionSeqCollection(
  monitors$pollution,
  summary_names = attr(monitors$pollution[[1]], "summary_names")
)
pm_trend <- dplyr::summarise(
  dplyr::group_by(enframe(pollution), event_seq),
  mean_pm = mean(Mean),
  median_pm = median(Median)
)

ggplot2::ggplot(pm_trend, ggplot2::aes(x = event_seq, y = mean_pm, group = 1)) +
  ggplot2::geom_line(color = "gray50") +
  ggplot2::geom_point(color = "palegreen4") +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "thistle4",
    linetype = "dashed"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = mean_pm),
    data = dplyr::filter(pm_trend, event_seq <= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = mean_pm),
    data = dplyr::filter(pm_trend, event_seq >= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(-12, 12, by = 2)
  ) +
  theme_minimal_light() +
  ggplot2::labs(
    x = "Months pre/post monitor start month",
    y = "Mean of de-seasonalized monitor mean PM2.5"
  )
```

Similarly, the median of monitor median de-seasonalized monthly values.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
ggplot2::ggplot(pm_trend, ggplot2::aes(x = event_seq, y = median_pm, group = 1)) +
  ggplot2::geom_line(color = "gray50") +
  ggplot2::geom_point(color = "palegreen4") +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "thistle4",
    linetype = "dashed"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = median_pm),
    data = dplyr::filter(pm_trend, event_seq <= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = median_pm),
    data = dplyr::filter(pm_trend, event_seq >= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(-12, 12, by = 2)
  ) +
  theme_minimal_light() +
  ggplot2::labs(
    x = "Months pre/post monitor start month",
    y = "Median of de-seasonalized monitor median PM2.5"
  )
```

### Log-transformed PM2.5 values

#### Non de-trended log(PM2.5) values

For each monitor, I've calculated the mean/median log(PM2.5) for the 12 months
before and after monitor installment. The mean of individual monitor means is
shown below.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
monitors <- readRDS(
  here::here("data/logged-pm/monitor-raw-pollution.Rds")
) |> dplyr::filter(row_id %in% all_monitors_dedup_site_buffer$row_id)
pollution <- new_pollutionSeqCollection(
  monitors$pollution,
  summary_names = attr(monitors$pollution[[1]], "summary_names")
)
pm_trend <- dplyr::summarise(
  dplyr::group_by(enframe(pollution), event_seq),
  mean_pm = mean(Mean),
  median_pm = median(Median)
)

ggplot2::ggplot(pm_trend, ggplot2::aes(x = event_seq, y = mean_pm, group = 1)) +
  ggplot2::geom_line(color = "gray50") +
  ggplot2::geom_point(color = "palegreen4") +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "thistle4",
    linetype = "dashed"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = mean_pm),
    data = dplyr::filter(pm_trend, event_seq <= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = mean_pm),
    data = dplyr::filter(pm_trend, event_seq >= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(-12, 12, by = 2)
  ) +
  theme_minimal_light() +
  ggplot2::labs(
    x = "Months pre/post monitor start month",
    y = "Mean of monitor mean log(PM2.5)"
  )
```

Similarly, the median of monitor median monthly values.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
ggplot2::ggplot(pm_trend, ggplot2::aes(x = event_seq, y = median_pm, group = 1)) +
  ggplot2::geom_line(color = "gray50") +
  ggplot2::geom_point(color = "palegreen4") +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "thistle4",
    linetype = "dashed"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = median_pm),
    data = dplyr::filter(pm_trend, event_seq <= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = median_pm),
    data = dplyr::filter(pm_trend, event_seq >= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(-12, 12, by = 2)
  ) +
  theme_minimal_light() +
  ggplot2::labs(
    x = "Months pre/post monitor start month",
    y = "Median of monitor median log(PM2.5)"
  )
```

#### log(PM2.5) values with seasonality removed (via STL)

Mean of de-seasonalized monitor means, but with log-transformed PM2.5 values.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
monitors <- readRDS(
  here::here("data/logged-pm/monitor-detrended-pollution.Rds")
) |> dplyr::filter(row_id %in% all_monitors_dedup_site_buffer$row_id)
pollution <- new_pollutionSeqCollection(
  monitors$pollution,
  summary_names = attr(monitors$pollution[[1]], "summary_names")
)
pm_trend <- dplyr::summarise(
  dplyr::group_by(enframe(pollution), event_seq),
  mean_pm = mean(Mean),
  median_pm = median(Median)
)

ggplot2::ggplot(pm_trend, ggplot2::aes(x = event_seq, y = mean_pm, group = 1)) +
  ggplot2::geom_line(color = "gray50") +
  ggplot2::geom_point(color = "palegreen4") +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "thistle4",
    linetype = "dashed"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = mean_pm),
    data = dplyr::filter(pm_trend, event_seq <= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = mean_pm),
    data = dplyr::filter(pm_trend, event_seq >= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(-12, 12, by = 2)
  ) +
  theme_minimal_light() +
  ggplot2::labs(
    x = "Months pre/post monitor start month",
    y = "Mean of de-seasonalized monitor mean log(PM2.5)"
  )
```

And finally, median of de-seasonalized median log(PM2.5) monitor values.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
ggplot2::ggplot(pm_trend, ggplot2::aes(x = event_seq, y = median_pm, group = 1)) +
  ggplot2::geom_line(color = "gray50") +
  ggplot2::geom_point(color = "palegreen4") +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "thistle4",
    linetype = "dashed"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = median_pm),
    data = dplyr::filter(pm_trend, event_seq <= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::geom_smooth(
    ggplot2::aes(x = event_seq, y = median_pm),
    data = dplyr::filter(pm_trend, event_seq >= 0),
    se = FALSE,
    method = "lm"
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(-12, 12, by = 2)
  ) +
  theme_minimal_light() +
  ggplot2::labs(
    x = "Months pre/post monitor start month",
    y = "Median of monitor median log(PM2.5)"
  )
```
